---
title: "1 Exploratory data analysis"
author: "stinekrye"
date: "2022-02-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
Exploratory data analysis

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(reshape2)
library(kableExtra)
library(ggpubr)
library(readxl)
library(gtools)
library(ggpointdensity)
library(viridis)
library(ggbeeswarm)
library(treemapify)

ASV_table <- read.delim("./data/oil_ASVtable.txt", sep = "\t", check.names = F)
sample_data <- read_excel("./data/sampleData.xlsx")
```


### Investigate the main problems found

**Count data** <br>
The total number of counts is limited by the sequencing depth, and can’t be compared directly. (Can't be shown by plots)

**Non-normally distributed data** <br>
Counts are not normally distributed. Neither are compositional data. This rules out t-test and MANOVA (Multivariate analysis of variance, (Anderson 2008))

Conclusion: The closest distribution I can tell is the log-normal distribution, and ~98% of this data set is zeroes!

#### First look at count data

```{r}
# Transform the data matrix to make it ready for plotting. Add log abundance variable
ASV_table.plot <- ASV_table %>% pivot_longer(cols = !ASVs, names_to = "sample_name", values_to = "abundance") %>% mutate(log_abundance = log10(abundance))

# Calculate basic statistics
stats <- ASV_table.plot %>% summarise_at(vars(abundance), list(Min = min, Mean = mean, Max = max, Sd = sd))

stats %>%  
  kable(caption="Statistics of count data") %>% 
  kable_styling()

# Calculate the percentage of zeroes in the data set
zeroes_percentage <- sum(ASV_table.plot$abundance == 0)/nrow(ASV_table.plot)
message(paste("The percentage of zeroes in the data set is:", round(zeroes_percentage,3), "%", collapse = ""))

# Remove all taxa with zero abundance
ASV_table.plot <- ASV_table.plot[ASV_table.plot$abundance > 0,]

```

A very large proportion of the data is zeroes. By removing those it is possible to see the pattern of all other taxa abundance:
```{r}
ASV_table.plot %>% ggplot(mapping = aes(x = abundance)) +
  geom_histogram(binwidth = 1000)+
  ylim(0,100)+
  ylab("Count")+
  xlab("Abundance ()")
```

```{r}
ASV_table.plot %>% gghistogram(x = "abundance",
   add = "mean", rug = TRUE, color = "aquamarine3", fill = "aquamarine2", bins= 15, xlab = "Abundance of individual taxa", ylab = "Counts")
```


Plot with log_abundance. This transformation makes it look more like a normal distribution, but it still has a large right tail.
```{r}
ASV_table.plot %>% ggdensity(x = "log_abundance",
   add = "mean", color = "aquamarine3", fill = "aquamarine2", bins= 30, xlab = "Log10 abundance of individual taxa", ylab = "Density")
```

#### Then look at compositional data

```{r}
ASV_table.plot <- ASV_table.plot %>% group_by(sample_name) %>% mutate(sample_sum = sum(abundance),
                                                            rel_abundance = abundance/sum(abundance),
                                                            log_rel_abundance = log10(rel_abundance)) %>% ungroup()

# Remove samples with sum lower than 20
ASV_table.plot <- ASV_table.plot[ASV_table.plot$sample_sum > 20,]

stats <- ASV_table.plot %>% summarise_at(vars(rel_abundance), list(Min = min, Mean = mean, Max = max, Sd = sd))
dominant <- ASV_table.plot[ASV_table.plot$rel_abundance > 0.8,]


stats %>%  
  kable(caption="Statistics of relative abundance") %>% 
  kable_styling()

```


The data looks approx. log-normal distributed

```{r}
ASV_table.plot %>% ggplot(mapping = aes(x = rel_abundance)) +
  geom_histogram(binwidth = 0.0001)+
  xlim(0.000001,0.5)+
  ylim(0,13000)+
  NULL
```

After a log transformation the data looks more normal, but does still deviate from (log) normal at very large and small relative abundances.

```{r}

ASV_table.plot %>% ggplot(mapping = aes(x = log10(rel_abundance))) +
  geom_histogram(binwidth = 0.1)
```

```{r}
ASV_table.plot %>% ggdensity(x = "log_rel_abundance",
   add = "mean", color = "aquamarine3", fill = "aquamarine2", bins= 15, xlab = "Log 10 relative abundance of individual taxa", ylab = "Density")
```

Normality tests

Log absolute abundance
```{r}
ggqqplot(ASV_table.plot$log_abundance) %>% ggpar(main = "Log10 abundance")
```

Log relative abundance
```{r}
ggqqplot(ASV_table.plot$log_rel_abundance) %>% ggpar(main = "Log10 relative abundance")
```


** Heteroscedasticity over a large dynamic range: **  <br>
(Love, Huber, and Anders 2014) Can’t use MANOVA and T-test. At least we need some transformations. In general I think it is hard to detect any difference in data if the dispersion is different AND the data does not follow a distribution.

IMPORTANT NOTE! All ASVs which has zero counts has been removed already!
Here I remove all ASVs which is NOT present in both of the replicates! This is a HUGE part of the remaining ASVs!

```{r}
#Add sampleName variable to data
sample_data <- sample_data %>% select(Samples, sampleName)
ASV_table.plot <- merge(x = ASV_table.plot, y = sample_data, by.x = "sample_name", by.y = "Samples") %>%  select(ASVs,log_abundance, abundance, rel_abundance, sampleName)

# Calculate the percentage of zeroes in the data set already removed
message(paste("The percentage of zeroes in the data set is:", round(zeroes_percentage,3), "%", collapse = ""))


total_ASVs <- nrow(ASV_table.plot)
removed_ASVs <- ASV_table.plot %>% group_by(ASVs, sampleName) %>% filter(n() < 2)
percentage_removed <- nrow(removed_ASVs)/total_ASVs

message(paste("Of the remaining part of the data set :", round(percentage_removed,3), "% are not present in both replicates and are removed", collapse = ""))

message(paste("Now we are left with", round((1-zeroes_percentage)*(1-percentage_removed),4), "% of the original data set", collapse = ""))
```


```{r}
# Caluclate mean of each ASV in each sample
# Remove ASVs from the sample if they are only found once in that sample
ASV_table.plot <- ASV_table.plot %>% group_by(ASVs, sampleName) %>% filter(n() >= 2) %>% ungroup()

# Calculate mean and sd for each ASV in each sample
ASV_table.plot <- ASV_table.plot %>% group_by(ASVs, sampleName) %>% mutate(abs_mean = mean(abundance),
                                                                           abs_sd = sd(abundance),
                                                                           rel_mean = mean(rel_abundance),
                                                                           rel_sd = sd(rel_abundance)) %>% ungroup()

message("Absolute abundance and relative abundance show the same pattern.")
ASV_table.plot %>%  ggplot(mapping = aes(x = abs_mean, y = abs_sd))+
  geom_pointdensity()+
  scale_color_viridis(option = "F")+
  geom_smooth(method = loess, color = "seagreen3")+
  geom_smooth(method = lm, color = "cornflowerblue")+
  stat_cor(method = "pearson", label.x = 0, label.y = 40000, color = "black")+
  ylab(expression("Standard deviation (relative abundance)"))+
  xlab(expression("Mean (relative abundance)"))+
  guides(color=guide_legend(title="Neighbors"))+
  theme_classic()
```

How to interpret these plot?

```{r}
message(" By dividing the standard deviation with the mean, we get this pattern")
ASV_table.plot %>%  ggplot(mapping = aes(x = abs_mean, y = abs_sd/abs_mean))+
  geom_pointdensity()+
  scale_color_viridis(option = "F")+
  geom_smooth(method = loess, color = "seagreen3")+
  geom_smooth(method = lm, color = "cornflowerblue")+
  stat_cor(method = "pearson", label.x = 10000, label.y = 2, color = "black")+
  # ylab(expression("Standard deviation (RA"^logit*")"))+
  # xlab(expression("Mean (RA"^logit*")"))+
  guides(color=guide_legend(title="Neighbors"))
```

Conclusion: The higher the variation is proportional to the size of the ASV!

Transformations which might fix this:

```{r}
# Transformations of the relative abundance
ASV_table.plot <- ASV_table.plot %>%  mutate(log_rel_abundance = log10(rel_abundance),
                                   log_it_rel_abundance = logit(rel_abundance, min = 0, max = 1)) %>% ungroup() # Add little constant?

# Calculate mean and sd
ASV_table.plot <- ASV_table.plot %>% group_by(ASVs, sampleName) %>% mutate(mean_log_rel_abundance = mean(log_rel_abundance),
                                                                 sd_log_rel_abundance = sd(log_rel_abundance),
                                                                 mean_logit_rel_abundance = mean(log_it_rel_abundance),
                                                                 sd_logit_rel_abundance = sd(log_it_rel_abundance)) %>% ungroup()

```


Logarithmic transformation

```{r}
subset <- ASV_table.plot[1:1000,]
```


```{r}
ASV_table.plot %>%  ggplot(mapping = aes(x = mean_log_rel_abundance, y = sd_log_rel_abundance))+
  geom_pointdensity()+
  scale_color_viridis(option = "F")+
  geom_smooth(method = loess, color = "seagreen3")+
  geom_smooth(method = lm, color = "cornflowerblue")+
  stat_cor(method = "pearson", label.x = -5, label.y = 2, color = "black")+
  guides(color=guide_legend(title="Neighbors"))+
  ylab(expression("Standard deviation (log10 relative abundance)"))+
  xlab(expression("Mean (log10 relative abundance)"))+
  scale_x_continuous(sec.axis = sec_axis(~10^.*100, name = "Relative abundance (%)", breaks = c(0.001, 0.003, 0.01, 0.03,  0.1, 0.3, 1, 3, 10, 30, 50, 80), labels=c("0.001", "0.003", "0.01", "0.03",  "0.1", "0.3", "1", "3", "10", "30", "50", "80")))+
  theme_classic()
```


Log-it transformation
```{r}
ASV_table.plot %>%  ggplot(mapping = aes(x = mean_logit_rel_abundance, y = sd_logit_rel_abundance))+
  geom_pointdensity()+
  scale_color_viridis(option = "F")+
  geom_smooth(method = loess, color = "seagreen3")+
  geom_smooth(method = lm, color = "cornflowerblue")+
  stat_cor(method = "pearson", label.x = -11, label.y = 5, color = "black")+
  guides(color=guide_legend(title="Neighbors"))+
  scale_x_continuous(sec.axis = sec_axis(~inv.logit(.)*100, name = "Relative abundance (%)", breaks = c(0.003, 0.01, 0.03,  0.1, 0.3, 1, 3, 10, 30, 50, 80), labels=c("0.003", "0.01", "0.03",  "0.1", "0.3", "1", "3", "10", "30", "50", "80")))+
  ylab(expression("Standard deviation (logit relative abundance)"))+
  xlab(expression("Mean (logit relative abundance)"))+
  theme_classic()
```



Logit relative abundance
```{r}
ggqqplot(ASV_table.plot$log_it_rel_abundance) %>% ggpar(main = "Logit relative abundance")

ASV_table.plot %>% ggdensity(x = "log_it_rel_abundance",
   add = "mean", color = "aquamarine3", fill = "aquamarine2", bins= 15, xlab = "Logit relative abundance of individual taxa", ylab = "Density")
```




** Outliers: ** <br>
Can drag the estimates/models
- It is hard to tell outliers apart from the rest of the data in this huge data set. I do not think this needs explannation.
- From the plots it is clear that the sd diverges around the very small / very large ASV sizes


** Number of taxa outnumber the samples: ** <br>
(Anderson 2008)
```{r}
dim_ASV <- dim(ASV_table)
message(paste("The dimension of the data is:", paste(dim_ASV, collapse = "/"), "(taxa/samples)", collapse = ""))
```

**Small replicate numbers** <br>
Is due to the cost of sequencing, which is fortunately decreasing. Results in lower power, uncertainty of within group variation. Pool samples (Love, Huber, and Anders 2014)
- It does not make sense to show this. I know that each group is only duplicates.



```{r}
composition <- data.frame("Category" = c("Zeroes", "Singletons", "Duplicates"), "Values" = c(round(zeroes_percentage,4), 
                                                                                             round((1-zeroes_percentage)*percentage_removed,4),
                                                                                             round((1-zeroes_percentage)-(1-zeroes_percentage)*percentage_removed,4)))


composition %>% ggplot(aes(area = Values, fill = Category)) +
  geom_treemap()
```


Block diagram for distribution of abundances (using bins)



# Estimate fold change


```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(matrixStats)

ASV_table <- read.delim("./data/oil_ASVtable.txt", sep = "\t", check.names = F)
```

```{r}
library(MatrixGenerics)
test <- ASV_table[1:10,] %>% select(-ASVs)
test[test == 0] <- NA
test <- as.matrix(test)
rowmed <- rowMedians(test, na.rm = T)
rowmax <- rowMaxs(test, na.rm = T)
diff <- rowmax/rowmed



test <- test %>% select(-ASVs)
test <- as.matrix(t(test))






colCV <- colSds(test)/colmeans(test)


test <-test %>% 
  mutate(min = min(across(everything())+1),
         max = max(across(everything())+1),
         Sum = sum(across(everything())),
         CV = sd/Mean) %>% arrange(desc(CV))


test %>% ggplot(aes(x))


test2 <- ASV_table[1:10,]
test %>% ggplot(aes(x = ASVs))

# test <- ASV_table %>% select(-ASVs) %>% mutate(sd = sd(across(everything())))
```




