---
title: "estimateSimulationParametersFromData"
author: "stinekrye"
date: "2022-04-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r Load libraries}
library(here)
here::i_am("./analysis/estimateSimulationParametersFromData.Rmd")
source(here("analysis", "sourceFunctions_compositionConstraintTests.R"))
source(here("analysis", "sourceFunctionsEstimateSimulationParametersFromData.R"))



library(tidyverse)
library(readxl)
library(phyloseq)
library(matrixStats)
library(vegan)
```

```{r Load data}
ASV_table <- read.delim("./data/oil_ASVtable.txt", sep = "\t", check.names = F, row.names = "ASVs")
sample_data <- read_excel("./data/sampleData.xlsx")
sample_data <- sample_data %>% column_to_rownames("Samples")
all_data <- phyloseq(otu_table(ASV_table, taxa_are_rows = T), sample_data(sample_data))
```

```{r Filter data}
# Extract GF data
GF <- subset_samples(all_data, region == "GF")

# Drop samples with sample siza < 10 and empty taxa
GF <- cleanData(GF)

# Extract DB data
DB <- subset_samples(all_data, region = "DB")

# Drop samples with sample siza < 10 and empty taxa
DB <- cleanData(DB)
```









? Drop technical replicates ?








```{r Estimate mean sample size and sd}
sampleSizesGF <- round(sample_sums(GF))
meanSampleSizeGF <- round(mean(sampleSizesGF))
medianSampleSizeGF <- round(median(sampleSizesGF))
meanSdSampleSizesGF <- round(sd(sampleSizesGF))

sampleSizesDB <- round(sample_sums(DB))
meanSampleSizeDB <- round(mean(sampleSizesDB))
medianSampleSizeDB <- ceiling(median(sampleSizesDB))
meanSdSampleSizesDB <- round(sd(sampleSizesDB))

hist(sampleSizesGF, breaks = 40)
hist(sampleSizesDB, breaks = 40, xlab = c("Sample size"), main = c("Disco Bay: Distribution of sample sizes")) # To make it look like the other. We 

d <- data.frame("x" = sampleSizesDB)

d %>% ggplot() +
  geom_histogram(aes(x = x), color = "black", fill = "firebrick")+
  xlab("Sample size")+
  ylab("Frequency")+
  ggtitle("Disko Bay: Distribution of sample sizes")+
  theme_classic()
```

```{r Estimate mean taxa size across all samples and sd}

GFotuTable <- as.matrix(as.data.frame(otu_table(GF)))
sdTaxaSizesGF <- round(rowSds(GFotuTable))
meanSdTaxaSizesGF <- round(mean(sdTaxaSizesGF))
meanTaxaSizeGF <- round(rowMeans(GFotuTable))
medianTaxaSizeGF <- round(rowMedians(GFotuTable))

DBotuTable <- as.matrix(as.data.frame(otu_table(DB)))
sdTaxaSizesDB <- round(rowSds(DBotuTable))
meanSdTaxaSizesDB <- round(mean(sdTaxaSizesDB))
meanTaxaSizeDB <- ceiling(rowMeans(DBotuTable))
medianTaxaSizeDB <- round(rowMedians(DBotuTable))

p <- density(log(meanTaxaSizeGF))
plot(p)
p <- density(log(meanTaxaSizeDB))
plot(p, xlim = range(-1,9), ylim = range(0,0.15))





d <- data.frame("x" = meanTaxaSizeDB)
# Mean across samples
d %>% ggplot() +
  geom_histogram(aes(x = x), color = "black", fill = "firebrick")+
  xlab("Mean taxa size ")+
  ylab("Frequency")+
  scale_x_continuous(trans = "log10")+
  ggtitle("Disko Bay: Mean taxa size")+
  theme_classic()
```

Taking the mean of taxa like this does not make biologial sense. DISCUSSION!



```{r Estimate dominant taxa in each sample}
# Highest abundance of a single taxon in each sample
maxTaxaGF <- apply(GFotuTable, MARGIN=c(2), max) 
maxTaxaDB <- apply(DBotuTable, MARGIN=c(2), max)


# Highest abundance divided with sample sum
domTaxaRatiosGF <- maxTaxaGF/sampleSizesGF
domTaxaRatiosDB <- maxTaxaDB/sampleSizesDB

# Look at max/mean/median domination
domMaxTaxaRatiosGF <- max(domTaxaRatiosGF)
domMinTaxaRatiosGF <- min(domTaxaRatiosGF)
domMedianTaxaRatiosGF <- median(domTaxaRatiosGF)
domMeanTaxaRatiosGF <- mean(domTaxaRatiosGF)

domMaxTaxaRatiosDB <- max(domTaxaRatiosDB)
domMinTaxaRatiosDB <- min(domTaxaRatiosDB)
domMedianTaxaRatiosDB <- median(domTaxaRatiosDB)
domMeanTaxaRatiosDB <- mean(domTaxaRatiosDB)

# Look at histograms to better understand how much of the sample is usually dominated by a single taxon
hist(domTaxaRatiosGF)
hist(domTaxaRatiosDB)


d <- data.frame("x" = domTaxaRatiosDB)

d %>% ggplot() +
  geom_histogram(aes(x = x), color = "black", fill = "firebrick")+
  xlab("Ratio of dominant taxa to total counts")+
  ylab("Frequency")+
  ggtitle("Disko Bay: Dominant taxa ratio")+
  theme_classic()
```
sampleSizesGF[maxTaxaGF/sampleSizesGF == 1]
Extend the above procedure to get an idea of how likely it is to observe domination by multiple species. How can I use this to adjust how the "diff abundance" step works?


? Zeroes ?

```{r}

GFzeroes <- apply(GFotuTable, MARGIN=c(2), function(x) sum(x == 0))
GFnonZeroes <- apply(GFotuTable, MARGIN=c(2), function(x) sum(x > 0))
GFzeroPercentage <- GFzeroes/(GFzeroes + GFnonZeroes)
hist(GFzeroPercentage)

DBzeroes <- apply(DBotuTable, MARGIN=c(2), function(x) sum(x == 0))
DBnonZeroes <- apply(DBotuTable, MARGIN=c(2), function(x) sum(x > 0))
DBzeroPercentage <- DBzeroes/(DBzeroes + DBnonZeroes)
hist(DBzeroPercentage)


d <- data.frame("x" = DBzeroPercentage)

d %>% ggplot() +
  geom_histogram(aes(x = x), color = "black", fill = "firebrick")+
  xlab("Percentage of zeroes")+
  ylab("Frequency")+
  ggtitle("Disko Bay: Zero percentage")+
  theme_classic()
```

```{r}
test <- rarecurve(t(DBotuTable), step = 500, label = F)
```


Idea: 
Average percentage in each sample?


<!-- # OLD Simulated data -->
<!-- ```{r Simulate data} -->
<!-- seed <- 128 -->
<!-- set.seed(seed) -->
<!-- change <- c(1) #c(1,2,5,8,15) -->
<!-- sample_size <-2# c(3,5,10,15,30) -->
<!-- runs <- 1 # 5 -->
<!-- n_taxa <- 8000#100 -->
<!-- n_random <- 0# c(2,5)#c(2,5)  -->
<!-- v <- 2 -->



<!-- diff_data_list <- data.frame() -->

<!-- for(k in seq(runs)){ -->
<!--   s <- k # Is set here, so the abundance is the same for taxon # in entire run k (So we can compare the effect of change + the other variables) -->
<!--   for(m in n_taxa) { -->
<!--     for(n in n_random) { -->
<!--       random_taxon <- sample(m, size = n, replace=F) # Randomly pick one taxon for differential abundance -->
<!--       for(j in sample_size){ -->
<!--         for(i in change){ -->
<!--           d <- createDfLogNormalAndNegBinom2(variables = v, sample_size = j, n_taxa = m, run = k, change = i, n_random = n,random_taxon = random_taxon, seed = s) -->
<!--           diff <- diffAbundance(df = d, change = i, random_taxon, variables = v) -->
<!--           diff_data_list <- bind_rows(diff_data_list, diff) -->
<!--         } -->
<!--       } -->
<!--     } -->
<!--   } -->
<!-- } -->


<!-- SIM <- diff_data_list %>% select(starts_with(c("1", "2"))) -->
<!-- SIM <- phyloseq(otu_table(SIM, taxa_are_rows = T)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- SIM <- cleanData(SIM) -->
<!-- ``` -->


<!-- ```{r Estimate mean sample size and sd} -->
<!-- sampleSizesSIM <- round(sample_sums(SIM)) -->
<!-- meanSampleSizeSIM <- round(mean(sampleSizesSIM)) -->
<!-- medianSampleSizeSIM <- round(median(sampleSizesSIM)) -->
<!-- meanSdSampleSizesSIM <- round(sd(sampleSizesSIM)) -->

<!-- hist(sampleSizesSIM) -->
<!-- ``` -->

<!-- ```{r Estimate mean taxa size across all samples and sd} -->

<!-- SIMotuTable <- as.matrix(as.data.frame(otu_table(SIM))) -->
<!-- sdTaxaSizesSIM <- round(rowSds(SIMotuTable)) -->
<!-- meanSdTaxaSizesSIM <- round(mean(sdTaxaSizesSIM)) -->
<!-- meanTaxaSizeSIM <- round(rowMeans(SIMotuTable)) -->
<!-- medianTaxaSizeSIM <- round(rowMedians(SIMotuTable)) -->

<!-- p <- density(log(meanTaxaSizeSIM)) -->
<!-- plot(p) -->
<!-- ``` -->

<!-- Taking the mean of taxa like this does not make biologial sense. DISCUSSION! -->



<!-- ```{r Estimate dominant taxa in each sample} -->
<!-- # Highest abundance of a single taxon in each sample -->
<!-- maxTaxaSIM <- apply(SIMotuTable, MARGIN=c(2), max)  -->


<!-- # Highest abundance divided with sample sum -->
<!-- domTaxaRatiosSIM <- maxTaxaSIM/sampleSizesSIM -->

<!-- # Look at max/mean/median domination -->
<!-- domMaxTaxaRatiosSIM <- max(domTaxaRatiosSIM) -->
<!-- domMinTaxaRatiosSIM <- min(domTaxaRatiosSIM) -->
<!-- domMedianTaxaRatiosSIM <- median(domTaxaRatiosSIM) -->
<!-- domMeanTaxaRatiosSIM <- mean(domTaxaRatiosSIM) -->

<!-- # Look at histograms to better understand how much of the sample is usually dominated by a single taxon -->
<!-- hist(domTaxaRatiosSIM) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- SIMzeroes <- apply(SIMotuTable, MARGIN=c(2), function(x) sum(x == 0)) -->
<!-- SIMnonZeroes <- apply(SIMotuTable, MARGIN=c(2), function(x) sum(x > 0)) -->
<!-- SIMzeroPercentage <- SIMzeroes/(SIMzeroes + SIMnonZeroes) -->
<!-- hist(SIMzeroPercentage) -->
<!-- ``` -->







# NEW Simulated data
```{r Simulate data}
source(here("analysis", "sourceFunctions_compositionConstraintTests.R"))
source(here("analysis", "sourceFunctionsEstimateSimulationParametersFromData.R"))
seed <- 120
set.seed(seed)
change <- c(10) #c(1,2,5,8,15)
sample_size <-c(150)# c(3,5,10,15,30)
runs <- 1 # 5
n_taxa <- 10000#100
n_random <- c(150) #2# c(2,5)#c(2,5) 
v <- 2



diff_data_list <- data.frame()

for(k in seq(runs)){
  s <- k # Is set here, so the abundance is the same for taxon # in entire run k (So we can compare the effect of change + the other variables)
  for(m in n_taxa) {
    for(n in n_random) {
      random_taxon <- sample(m, size = n, replace=F) # Randomly pick one taxon for differential abundance
      for(j in sample_size){
        for(i in change){
          d <- createDfLogNormalAndNegBinom3(variables = v, sample_size = j, n_taxa = m, run = k, change = i, n_random = n, seed = s)
          diff <- diffAbundance(df = d, change = i, random_taxon, variables = v)
          diff_data_list <- bind_rows(diff_data_list, diff)
        }
      }
    }
  }
}

diff_data_list <- diff_data_list %>% select(-starts_with("V"))
SIM <- diff_data_list %>% select(starts_with(c("1", "2")))
SIM <- phyloseq(otu_table(SIM, taxa_are_rows = T))
```

```{r}
SIM <- cleanData(SIM)
```

```{r}
# t <- SIMotuTable[,1:2]
# test <- rarecurve(t(SIMotuTable), step = 500, label = F)
```


```{r Estimate mean sample size and sd}
sampleSizesSIM <- round(sample_sums(SIM))
meanSampleSizeSIM <- round(mean(sampleSizesSIM))
medianSampleSizeSIM <- round(median(sampleSizesSIM))
meanSdSampleSizesSIM <- round(sd(sampleSizesSIM))

hist(sampleSizesSIM, breaks = 10)

d <- data.frame("x" = sampleSizesSIM)

d %>% ggplot() +
  geom_histogram(aes(x = x), color = "black", fill = "darkgoldenrod3")+
  xlab("Sample size")+
  ylab("Frequency")+
  ggtitle("Simulated data: Distribution of sample sizes")+
  scale_x_continuous(n.breaks = 5, limits = c(19000,40000))+
  # xlim(20550,20650)+
  theme_classic()
```


```{r Estimate mean taxa size across all samples and sd}

SIMotuTable <- as.matrix(as.data.frame(otu_table(SIM)))
sdTaxaSizesSIM <- round(rowSds(SIMotuTable))
meanSdTaxaSizesSIM <- round(mean(sdTaxaSizesSIM))
meanTaxaSizeSIM <- round(rowMeans(SIMotuTable))
medianTaxaSizeSIM <- round(rowMedians(SIMotuTable))

p <- density(log(meanTaxaSizeSIM), bw = 0.2484)
plot(p)

d <- data.frame("x" = meanTaxaSizeSIM)
# Mean across samples
d %>% ggplot() +
  geom_histogram(aes(x = x), color = "black", fill = "darkgoldenrod3")+
  xlab("Mean taxa size ")+
  ylab("Frequency")+
  scale_x_continuous(trans = "log10")+
  ggtitle("Simulated data: Mean taxa size")+
  ylim(0,115)+
  theme_classic()

```

Taking the mean of taxa like this does not make biologial sense. DISCUSSION!



```{r Estimate dominant taxa in each sample}
# Highest abundance of a single taxon in each sample
maxTaxaSIM <- apply(SIMotuTable, MARGIN=c(2), max) 


# Highest abundance divided with sample sum
domTaxaRatiosSIM <- maxTaxaSIM/sampleSizesSIM

# Look at max/mean/median domination
domMaxTaxaRatiosSIM <- max(domTaxaRatiosSIM)
domMinTaxaRatiosSIM <- min(domTaxaRatiosSIM)
domMedianTaxaRatiosSIM <- median(domTaxaRatiosSIM)
domMeanTaxaRatiosSIM <- mean(domTaxaRatiosSIM)

# Look at histograms to better understand how much of the sample is usually dominated by a single taxon
hist(domTaxaRatiosSIM)

d <- data.frame("x" = domTaxaRatiosSIM)

d %>% ggplot() +
  geom_histogram(aes(x = x), color = "black", fill = "darkgoldenrod3")+
  xlab("Ratio of dominant taxa to total counts")+
  ylab("Frequency")+
  ggtitle("Simulated data: Dominant taxa ratio")+
  scale_x_continuous(n.breaks = 5, limits = c(0,1))+
  # ylim(0,50)+
  theme_classic()
```

```{r}
SIMzeroes <- apply(SIMotuTable, MARGIN=c(2), function(x) sum(x == 0))
SIMnonZeroes <- apply(SIMotuTable, MARGIN=c(2), function(x) sum(x > 0))
SIMzeroPercentage <- SIMzeroes/(SIMzeroes + SIMnonZeroes)
hist(SIMzeroPercentage)

d <- data.frame("x" = SIMzeroPercentage)

d %>% ggplot() +
  geom_histogram(aes(x = x), color = "black", fill = "darkgoldenrod3")+
  xlab("Percentage of zeroes")+
  ylab("Frequency")+
  ggtitle("Simulated data: Zero percentage")+
  scale_x_continuous(n.breaks = 5, limits = c(0.9,1))+
  # ylim(0,60)+
  # xlim(0.9,1)+
  theme_classic()
```


Dispersion figure

```{r}
# dataSim <- read.csv("./data/dataFinal/simulatedMergedClrnoClr/ConfLogNormalAndNegBinom3SampleSize3#preThresh(0.05).csv", check.names = F)

# Cut out the data you need
dataSimClean <- diff_data_list %>% select(-contains(as.character(seq(4,9)))) %>% distinct() %>% select("run","taxon", "2.1", "2.2", "2.3")

# Drop empty rows
dataSimCleanGroupOne <- dataSimClean[rowSums(dataSimClean[3:5]) > 0,]

# Control prevalence
prevalence <- 0.99
dataSimCleanGroupOne <- dataSimCleanGroupOne[rowSums(dataSimCleanGroupOne[3:5] > 0, na.rm = T)/ncol(dataSimCleanGroupOne[3:5]) >= prevalence,]
```

```{r}


# Reshape df
plotDataSimCleanGroupOne <- dataSimCleanGroupOne %>%  pivot_longer(cols = c(-taxon, -run), names_to = "sample", values_to = "count")

# Rel abundance and trans?
plotDataSimCleanGroupOne <- plotDataSimCleanGroupOne %>% group_by(run, sample) %>% mutate(relAbundance = count/sum(count),
                                                                      logitTrans = log(((relAbundance-0+0.000000001)/(1-0))/(1-(relAbundance-0)/(1-0))),
                                                                      logTrans = log10(relAbundance))

# Calculate mean and sd
plotDataSimCleanGroupOne <- plotDataSimCleanGroupOne %>% group_by(run,taxon) %>% summarise(meanRel = mean(relAbundance),
                                                                                           stddevRel = sd(relAbundance),
                                                                                           meanLogit = mean(logitTrans),
                                                                                           stddevLogit = sd(logitTrans),
                                                                                           meanLog = mean(logTrans),
                                                                                           stddevLog = sd(logTrans))

```


```{r}
testPlotDataSimCleanGroupOne <- plotDataSimCleanGroupOne %>% filter(run %in% c(1,2,3))

simP <- testPlotDataSimCleanGroupOne %>% ggplot(mapping = aes(x = meanLog, y = stddevLog))+ # V1 is mean and V2 is std dev
  geom_point(size = 1)+
  # geom_smooth(method = loess, color = "seagreen3", alpha = 0.1)+
  # geom_smooth(method = lm, color = "cornflowerblue", alpha = 0.1)+
  # stat_cor(method = "pearson", label.x = -1.5, label.y = 0.75, color = "black")+
  ylim(0,0.8)+
  ylab(expression("Standard deviation"))+
  xlab(expression("Mean (log"[10]*"(Relative abundance))"))+
  guides(color=guide_legend(title="Neighbors"))+
  scale_x_continuous(limits = c(-5,0),sec.axis = sec_axis(~10^(.)*100, name = "Relative abundance (%)", breaks = c(0.001, 0.005, 0.03,  0.1, 0.3, 1, 3, 10, 30, 50), labels=c("0.001","0.005", "0.03",  "0.1", "0.3", "1", "3", "10", "30", "50")))+
  theme_classic()+
  ggtitle("Dispersion of relative abundance of taxon in sim. data")+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=0.5))

simP
```



Real data

```{r}
dataReal <- read.csv("./data/dataFinal/realData/realDataFull#preThresh(0.05).csv", check.names = F)

# Cut out the data you need
dataRealClean <- dataReal %>% select(-contains(c(as.character(seq(4,9))))) %>% distinct() %>% select("taxon", "1.1", "1.2", "1.3")

# Drop empty rows
dataRealCleanGroupOne <- dataRealClean[rowSums(dataRealClean[2:4]) > 0,]

# Control prevalence
prevalence <- 0.99
dataRealCleanGroupOne <- dataRealCleanGroupOne[rowSums(dataRealCleanGroupOne[2:4] > 0, na.rm = T)/ncol(dataRealCleanGroupOne[2:4]) >= prevalence,]
```

```{r}


# Reshape df
plotDataRealCleanGroupOne <- dataRealCleanGroupOne %>%  pivot_longer(cols = c(-taxon), names_to = "sample", values_to = "count")

# Rel abundance and trans?
plotDataRealCleanGroupOne <- plotDataRealCleanGroupOne %>% group_by(sample) %>% mutate(relAbundance = count/sum(count),
                                                                      logitTrans = log(((relAbundance-0+0.000000001)/(1-0))/(1-(relAbundance-0)/(1-0))),
                                                                      logTrans = log10(relAbundance))

# Calculate mean and sd
plotDataRealCleanGroupOne <- plotDataRealCleanGroupOne %>% group_by(taxon) %>% summarise(meanRel = mean(relAbundance),
                                                                                           stddevRel = sd(relAbundance),
                                                                                           meanLogit = mean(logitTrans),
                                                                                           stddevLogit = sd(logitTrans),
                                                                                           meanLog = mean(logTrans),
                                                                                           stddevLog = sd(logTrans))

```


```{r}


realP <- plotDataRealCleanGroupOne %>% ggplot(mapping = aes(x = meanLog, y = stddevLog))+ # V1 is mean and V2 is std dev
  geom_point(size = 1)+
  # geom_smooth(method = loess, color = "seagreen3", alpha = 0.1)+
  # geom_smooth(method = lm, color = "cornflowerblue", alpha = 0.1)+
  # stat_cor(method = "pearson", label.x = -1.5, label.y = 0.75, color = "black")+
  ylim(0,0.8)+
  ylab(expression("Standard deviation"))+
  xlab(expression("Mean (log"[10]*"(Relative abundance))"))+
  guides(color=guide_legend(title="Neighbors"))+
  scale_x_continuous(limits = c(-5,0),sec.axis = sec_axis(~10^(.)*100, name = "Relative abundance (%)", breaks = c(0.001, 0.005, 0.03,  0.1, 0.3, 1, 3, 10, 30, 50), labels=c("0.001","0.005", "0.03",  "0.1", "0.3", "1", "3", "10", "30", "50")))+
  theme_classic()+
  ggtitle("Dispersion of relative abundance of taxon in real data")+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=0.5))

realP
```


