---
title: "5_realDataTest"
author: "stinekrye"
date: "2022-05-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r Load libraries}
library(here)
here::i_am("./analysis/5_realDataTest.Rmd")
source(here("analysis", "sourceFunctions_compositionConstraintTests.R"))
source(here("analysis", "sourceFunctionsRealDataTest.R"))

library(tidyverse)
library(readxl)
library(ggplot2)
library(phyloseq)
library(reshape2)
library(qvalue)
library(kableExtra)
library(DESeq2)
library(ALDEx2)
library(ANCOMBC)
library(rstatix)
library(parallel)
library(ggvenn)
library(devtools)
library(RColorBrewer)
library(venn)
library(cowplot)
library(viridis)
library(compositions)
library(vegan)
```

```{r}
theme_set(theme_bw())
pal =  "Set1"# "RdYlGn"# RdYIBu
scale_colour_discrete <-  function(palname=pal, ...){
  scale_colour_brewer(palette=palname, direction=-1, ...)
}
scale_fill_discrete <-  function(palname=pal, ...){
  scale_fill_brewer(palette=palname, direction=-1, ...)
}
```

```{r Load data}
ASV_table <- read.delim("./data/oil_ASVtable.txt", sep = "\t", check.names = F, row.names = "ASVs")
sample_data <- read_excel("./data/sampleData.xlsx")
sample_data <- sample_data %>% column_to_rownames("Samples")
sample_data <- sample_data %>% select(-c("timespan", "QPCRCopies", "QPCR_SD","Pressure (Mpa)", "phosphate", "nitrate", "nitrite", "DIC", "DIC_SD"))
sample_data <- sample_data %>% rename(days = "Days",
                                        location = "Location")
tree <-  read_tree("./data/oil_spill.nwk")
```

```{r Identify replicates and best subgroup (variable: type)}
# Save rownames
sample_data <- rownames_to_column(sample_data, var = "rowname")

# Save metadata from DB
# sample_data <- sample_data %>% filter(region == "DB")

# Assign a unique identifier to each group and count size of groups (Grouped by ALL variables)
sample_data <- sample_data %>% group_by(region, type, Location,lightDark, depth, Days, oil, nutrient, freeLivingBiofilm) %>% mutate(ID = cur_group_id(), n = n())

# Make column which tells which samples to keep (>=3 replicates)
sample_data <-  sample_data %>% mutate(keep = ifelse(n >= 3, "T", "F"))

# impute all Days < 40 with 37. In DB data Days = 34, 37 and 38 for three different Locations. I think it is ok to think this as the same
sample_data <- sample_data %>% mutate(Days = replace(Days, Days < 40, 36))

# impute all Days == 117 with 116. I think it is ok to think this as the same
sample_data <- sample_data %>% mutate(Days = replace(Days, Days == 117, 116))

#
sample_data$Days <- as.character(sample_data$Days)


# Keep only data with the number of replicates >= 3
test <- sample_data %>% filter(n >= 3)

# Find the type with most samples with three replicates
table(test$type) # insitu is largest
```

```{r Make phyloseq object and subset}
# Recreate rownames
sample_data <- column_to_rownames(sample_data, "rowname")

# Make phyloseq object with full data
data <- phyloseq(otu_table(ASV_table, taxa_are_rows = T), sample_data(sample_data), tree)

# Subset data
data <- subset_samples(data, region == "DB" & type == "insitu" & keep == "T")


# End up with 52 (Yes)
```

```{r Clean phyloseq data}
# Remove empty taxa
data <- prune_taxa(!taxa_sums(data) == 0, data)

# Look at sample size
sample_sums(data)

# Look at sample rarefaction curve
asvs <- t(otu_table(data))
test <- rarecurve(asvs, step = 1000, label = F) # Looks fine even for the small samples

# Remove FG-9_S128_F_filt.fastq.gz. Have been rerun as FG-9_S107. ID 69
dataClean <- prune_samples(!sample_names(data) == "FG-9_S128_F_filt.fastq.gz", data)

# Transform data to relative abundance
dataCleanRel <- transform_sample_counts(dataClean, function(x) (100 * x/sum(x) ) )

# Perform clr transformation
dataCleanClr <- transform_sample_counts(dataCleanRel, clr) # IS SAMPLEWISE!!!!! Hurray!

# Perform log trans
dataCleanLog <- transform_sample_counts(dataClean, log) # IS SAMPLEWISE!!!!! Hurray!
```

Notes:
- The variables in the data are now: Location, Days and oil


```{r Explore data: Alpha diversity - Plot 1}

p <- plot_richness(dataCleanClr, x = "Days" , color = "oil", measures = c("InvSimpson"))
# p <- p + geom_line()
p <- p + guides(color=guide_legend(title="Oil type"))
p <- p + facet_wrap("Location")
# p <- p + scale_x_continuous(breaks = c(1,2,3,4), labels= c("8", "37", "100", "379"))
p <- p + xlab("Time (Days)")
p <- p + ylab(" Alpha Diversity (Inverse Simpson Index)")
p <- p + stat_summary(fun=mean, geom="line")
# p <- p + scale_y_log10()
# p$layers[[1]] <- NULL
p

```

```{r Explore data: Beta diversity - Plot 1}
dataCleanClr.ord <- ordinate(dataCleanClr, method = "NMDS", distance = "unifrac") #jaccard, chao, binomial, chord, wunifrac, gower, jsd # k3

my_colors <- brewer.pal(9,"Spectral")[c(1,3,5,7,9)]
p1 = plot_ordination(dataCleanClr, dataCleanClr.ord, type="samples", title="Disko Bay NMDS ordination")
p1 <- p1 + facet_wrap("Location", labeller = label_both)
p1 <- p1 + scale_shape_manual(values = c(21,22,24)) #name = "Location", labels = c("Deep", "Sediment", "Surface/Dark", "Surface/Light"))
# p1 <- p1 + scale_color_manual(values = my_colors, name = "Days")
p1 <- p1 + geom_point(aes(fill = Days, shape = oil), color = "black", size = 3)
p1 <- p1 + guides(fill = guide_legend(title = "Days", override.aes = list(shape = 21)),
                  shape = guide_legend(title = "Oil Type"))
# guides(fill = guide_legend(override.aes = list(shape = 21)))
p1 <- p1 + theme_classic()
print(p1)
```

```{r Explore data: Beta diversity - Plot 2}
dataCleanClr.ord <- ordinate(dataCleanClr, method = "NMDS", distance = "unifrac") #jaccard, chao, binomial, chord, wunifrac, gower, jsd # k3

my_colors <- brewer.pal(9,"Spectral")[c(1,3,5,7,9)]
p2 = plot_ordination(dataCleanClr, dataCleanClr.ord, type="samples", title="Disko Bay NMDS ordination")
p2 <- p2 + facet_wrap("Days",labeller = label_both)
p2 <- p2 + scale_shape_manual(values = c(21,22,24)) #name = "Location", labels = c("Deep", "Sediment", "Surface/Dark", "Surface/Light"))
# p2 <- p2 + scale_color_manual(values = my_colors, name = "Days")
p2 <- p2 + geom_point(aes(fill = Location, shape = oil), color = "black", size = 3)
p2 <- p2 + guides(fill = guide_legend(title = "Location", override.aes = list(shape = 21)),
                  shape = guide_legend(title = "Oil Type"))
# guides(fill = guide_legend(override.aes = list(shape = 21)))
p2 <- p2 + theme_classic()
print(p2)
```



Calculate distances and make test
```{r Explore data: Beta diversity - PERMANOVA}
do_adonis <- function(phylo, subset){
dist <- UniFrac(phylo)
metadata <- data.frame(sample_data(phylo))
if(subset == TRUE) {
  test <- adonis2(dist ~ Location + oil, data = metadata, permutations = 9999, by = "margin")
  return(test)
}
test <- adonis2(dist ~ Location + Days + oil, data = metadata, permutations = 9999, by = "margin")
return(test)

}
```

Full dataset
```{r}
alltime_adonis <- do_adonis(dataCleanClr, subset = F)
```



Subsample by timepoint
```{r}
onemonth <- subset_samples(dataCleanClr,  Days == 36)
threemonths <- subset_samples(dataCleanClr,  Days == 116)



onemonth_adonis <- do_adonis(onemonth, subset = T)
threemonths_adonis <- do_adonis(threemonths, subset = T)

```



Noter til beta diversity plot 1:
- Tid + lokation skiller prøverne helt ad. Hvilke organismer ændrer sig over tid?
  - Test: Grupper data efter lokation, olie og tid. Sammenlign de to tider indenfor hver lokation. Udvælg evt. to grupper, der er særligt godt adskilt. F.eks. trekanterne i lokation C. VARIABEL SOM TESTES : TID
  - Test 2: Grupper data efter lokation, olie og tid. Sammenlign de blå cirkler med enten de blå trekanter eller firkanter. VARIABEL SOM TESTES: OLIE (TREATMENT)

Noter til beta diversity plot 2:
- Lokation er god til at adskille data ved dag 36, men knapt så god ved dag 116. 
  - Test: Sammenlign de farvede trekanter ved dag 36. Hvilke taxa er forskellige?




Q: Is it a problem that the root in the three is reassigned each time?
Q: Technical or biological replicates?

# Differential abundance analysis

```{r}
options(pillar.sigfig=8)
data <- dataClean
source(here("analysis", "sourceFunctionsRealDataTest.R"))
seed <- 120
prevalence <- 1
tests <- c("DESEq2", "Wilcoxon", "Student.t.test", "ALDEx2", "ANCOM.BC") #"DESEq2", "Wilcoxon", "Student.t.test", "ALDEx2", "ANCOM.BC"
data <- prepareData(data)
data <- data[-7] # Manually remove data frames which are note complete (missing a sample)
```


```{r}
test <- runAllReal(data, tests, prevalence, seed)
write_csv(test, "realDataFull#preThresh(1).csv")
```

```{r}
# Redo with duplicates

test <- runAllRealDup(data, tests, prevalence, seed, c("2","3"))
# write_csv(test, "realData(2.3)#preThresh(1).csv")
```



